\documentclass{thesis}
\begin{document}
\pagenumbering{gobble}
\chapter*{Acknowledgements}
I am immensely thankful to all the people who have supported me throughout my PhD.

First and foremost, a big thank you to my supervisor Ewan Birney who gave me the opportunity to work in his research group. Ewan's enthusiasm, ideas and support were invaluable to guide me through the past four years. I would also like to thank the other members --current and past-- of the Birney research group: Ian Dunham, Sander Timmer, Sandro Morganella, Valentina Iotchkova (thank you for all your advice on the statistics), Helena Kilpinen, Leland Taylor, Tom Fitzgerald , Carl Barton, Anat Melamed and the other Hannah (Currant). I really valued the helpful discussions and feedback and enjoyed our times in the Northumbrian wilderness. A special thanks to Tracey Andrews, Stacy Knoop, Debbie Howe and Christina Karikidis for their help with all the administrative tasks and finding time in Ewan's busy schedule to meet with me.

My PhD project would not have been possible without my collaboration partners. Paolo Casale and Oliver Stegle supported me with their experience and constructive advice in the method development part of this thesis. Konrad Rudolph helped with the development of the R package. The heart project was based on a close collaboration with Stuart Cook's research group in London, in particular with Antonio De Marvao and Declan O'Regan. My thesis advisory committee with John Marioni, Carl Anderson and Jan Korbel was very helpful with their constructive criticism and suggestions, keeping me and Ewan on track and helped with the timely submission of this thesis. 

Many thanks to the people who helped with proofreading of this thesis: the two Nils (Eling and K\"olling), Jack Monagan, Hannah Currant, Carl Barton, Paolo Casale and David Pattinson (finding the biggest typo in a chapter heading on the day of submission).

Aside from the academic support, there are many people who deserve a special thank you for their moral support and friendship over the past years. My experience would not have been the same without the countless hours of early mornings on the river with Jesus College Boat Club, the exciting volleyball matches with the Cambridge University Volleyball Club and the great people I met in both teams. I would like to thank my fellow EMBL and Cambridge PhD students for the fun times during the predocs course, the many Cambridge formals and warm welcomes when coming back to Heidelberg, especially Nils, Silvia, Christina, Julia, Laura, Joran, Kostek, Chris, Paul and my housemates Maria and Dani. To my close friends from afar, Micha, Melanie, Jana, Lissy, Geli, Leonie and Meike (not quite so far): your support and encouragement, at whatever time and place was invaluable. I would like to thank Dave, for the discussions and input on my research, the outdoors activities to distract from them, but mostly, for everything else. 

Finally, I would like to thank my family who supported me in all my decisions, encouraged me in difficult times and shared my happiness in good ones. 

\end{document}
